---
title: "Scammy pins and power!"
format: html
editor: visual
---

```{r}
#| label: helper-functions
#| code-fold: true
#| code-summary: "Show helper functions"

# function for difference of proportions
plot_diff_prop_dist <- function(n1, n2, p1, p2,
                                hyp = "the null hypothesis",
                                lower_cut = NULL,
                                upper_cut = NULL,
                                obs_T = NULL,
                                two_tailed = FALSE,
                                show_tails = TRUE,
                                mode = c("pvalue", "power"),
                                lwd = 2,
                                tail_col = NULL,
                                body_col = "black",
                                pch = 19,
                                add = FALSE,
                                xlim = NULL,
                                ylim = NULL,
                                tol_snap = 1e-8) {
  
  mode <- match.arg(mode)
  
  if (is.null(tail_col)) {
    tail_col <- if (mode == "power") "blue" else "red"
  }
  
  # exact distribution
  x1 <- 0:n1
  x2 <- 0:n2
  
  prob1 <- dbinom(x1, size = n1, prob = p1)
  prob2 <- dbinom(x2, size = n2, prob = p2)
  
  grid <- expand.grid(x1 = x1, x2 = x2)
  grid$Tval <- round(grid$x2 / n2 - grid$x1 / n1, 12)
  grid$prob <- prob1[grid$x1 + 1] * prob2[grid$x2 + 1]
  
  dist_df <- aggregate(prob ~ Tval, data = grid, sum)
  dist_df <- dist_df[order(dist_df$Tval), ]
  
  Tvals <- dist_df$Tval
  probs <- dist_df$prob
  
  lower_tail <- rep(FALSE, length(Tvals))
  upper_tail <- rep(FALSE, length(Tvals))
  
  if (mode == "pvalue") {
    
    if (two_tailed) {
      if (is.null(obs_T)) {
        stop("If two_tailed = TRUE, you must provide obs_T.")
      }
      
      j_obs <- which.min(abs(Tvals - obs_T))
      if (abs(Tvals[j_obs] - obs_T) > tol_snap) {
        stop("obs_T is not close enough to an attainable value.")
      }
      obs_T <- Tvals[j_obs]
      
      center <- p2 - p1
      
      if (obs_T <= center) {
        lower_tail <- Tvals <= obs_T
        lower_prob <- sum(probs[lower_tail])
        
        cum_prob <- 0
        for (i in length(Tvals):1) {
          if (Tvals[i] > center && cum_prob + probs[i] <= lower_prob + tol_snap) {
            upper_tail[i] <- TRUE
            cum_prob <- cum_prob + probs[i]
          }
        }
        upper_prob <- sum(probs[upper_tail])
      } else {
        upper_tail <- Tvals >= obs_T
        upper_prob <- sum(probs[upper_tail])
        
        cum_prob <- 0
        for (i in seq_along(Tvals)) {
          if (Tvals[i] < center && cum_prob + probs[i] <= upper_prob + tol_snap) {
            lower_tail[i] <- TRUE
            cum_prob <- cum_prob + probs[i]
          }
        }
        lower_prob <- sum(probs[lower_tail])
      }
      
      total_tail <- lower_prob + upper_prob
      
      main_title <- bquote(
        atop(
          "Exact sampling distribution of " ~
            T["(n"[1]*","*n[2]*")"] == hat(p)[2]["(n"[2]*")"] - hat(p)[1]["(n"[1]*")"],
          "under " ~ .(hyp) ~ "," ~
            p[1] == .(p1) ~ "," ~ p[2] == .(p2) ~ "," ~
            n[1] == .(n1) ~ "," ~ n[2] == .(n2)
        )
      )
      
      sub_title <- bquote(
        "Observed =" ~ .(obs_T) ~ "," ~
        "left tail =" ~ .(round(lower_prob, 4)) ~ "," ~
        "right tail =" ~ .(round(upper_prob, 4)) ~ "," ~
        "two-sided p-value =" ~ .(round(total_tail, 4))
      )
      
    } else {
      if (!is.null(lower_cut)) lower_tail <- Tvals <= lower_cut
      if (!is.null(upper_cut)) upper_tail <- Tvals >= upper_cut
      
      lower_prob <- sum(probs[lower_tail])
      upper_prob <- sum(probs[upper_tail])
      total_tail <- lower_prob + upper_prob
      
      main_title <- bquote(
        atop(
          "Exact sampling distribution of " ~
            T["(n"[1]*","*n[2]*")"] == hat(p)[2]["(n"[2]*")"] - hat(p)[1]["(n"[1]*")"],
          "under " ~ .(hyp) ~ "," ~
            p[1] == .(p1) ~ "," ~ p[2] == .(p2) ~ "," ~
            n[1] == .(n1) ~ "," ~ n[2] == .(n2)
        )
      )
      
      sub_title <- NULL
      if (!is.null(lower_cut) || !is.null(upper_cut)) {
        sub_title <- bquote(
          "Lower tail =" ~ .(round(lower_prob, 4)) ~ "," ~
          "Upper tail =" ~ .(round(upper_prob, 4)) ~ "," ~
          "Total p-value =" ~ .(round(total_tail, 4))
        )
      }
    }
  }
  
  if (mode == "power") {
    if (is.null(lower_cut) && is.null(upper_cut)) {
      stop("In power mode, provide at least one of lower_cut or upper_cut.")
    }
    
    if (!is.null(lower_cut)) lower_tail <- Tvals <= lower_cut
    if (!is.null(upper_cut)) upper_tail <- Tvals >= upper_cut
    
    lower_prob <- sum(probs[lower_tail])
    upper_prob <- sum(probs[upper_tail])
    total_tail <- lower_prob + upper_prob
    
    main_title <- bquote(
      atop(
        "Exact sampling distribution of " ~
          T["(n"[1]*","*n[2]*")"] == hat(p)[2]["(n"[2]*")"] - hat(p)[1]["(n"[1]*")"],
        "under the following alternative distribution," ~
          p[1] == .(p1) ~ "," ~ p[2] == .(p2) ~ "," ~
          n[1] == .(n1) ~ "," ~ n[2] == .(n2)
      )
    )
    
    sub_title <- bquote(
      "Probability to reject null =" ~ .(round(total_tail, 4)) ~
      "=" ~ "power with respect to" ~ .(hyp)
    )
  }
  
  tail_region <- lower_tail | upper_tail
  cols <- if (show_tails) ifelse(tail_region, tail_col, body_col) else rep(body_col, length(Tvals))
  
if (is.null(xlim)) xlim <- range(Tvals)
if (is.null(ylim)) ylim <- c(0, 1.1 * max(probs))

plot(Tvals, probs,
     type = "h",
     lwd = lwd,
     col = cols,
     xlab = expression(hat(p)[2]["(n"[2]*")"] - hat(p)[1]["(n"[1]*")"]),
     ylab = "Probability",
     main = main_title,
     sub = sub_title,
     xlim = xlim,
     ylim = ylim)

points(Tvals, probs, col = cols, pch = pch)
  
  invisible(list(
    Tvals = Tvals,
    probs = probs,
    lower_tail = lower_tail,
    upper_tail = upper_tail,
    total = total_tail
  ))
}
```

```{r}

#| label: helper-functions
#| code-fold: true
#| code-summary: "Show helper functions"
#| 
plot_diff_prop_normal <- function(n1, n2, p1, p2,
                                  hyp = "the null hypothesis",
                                  lower_cut = NULL,
                                  upper_cut = NULL,
                                  obs_T = NULL,
                                  two_tailed = FALSE,
                                  show_tails = TRUE,
                                  mode = c("pvalue", "power"),
                                  lwd = 2,
                                  tail_col = NULL,
                                  body_col = "black",
                                  ngrid = 2000,
                                  add = FALSE,
                                  xlim = NULL,
                                  ylim = NULL) {
  
  mode <- match.arg(mode)
  
  if (is.null(tail_col)) {
    if (mode == "pvalue") tail_col <- "red"
    if (mode == "power")  tail_col <- "blue"
  }
  
  mu <- p2 - p1
  sigma <- sqrt(p1 * (1 - p1) / n1 + p2 * (1 - p2) / n2)
  
  if (sigma == 0) {
    stop("The normal approximation is degenerate here because sigma = 0.")
  }
  
  x_min <- mu - 4 * sigma
  x_max <- mu + 4 * sigma
  x <- seq(x_min, x_max, length.out = ngrid)
  dens <- dnorm(x, mean = mu, sd = sigma)
  
  lower_prob <- 0
  upper_prob <- 0
  total_tail <- 0
  
  shade_left <- FALSE
  shade_right <- FALSE
  lower_bound <- NA
  upper_bound <- NA
  
  if (mode == "pvalue") {
    
    if (two_tailed) {
      if (is.null(obs_T)) {
        stop("If two_tailed = TRUE, you must provide obs_T.")
      }
      
      if (obs_T <= mu) {
        lower_bound <- obs_T
        lower_prob <- pnorm(obs_T, mean = mu, sd = sigma)
        
        # opposite-side cutoff chosen so right tail has same probability
        upper_bound <- qnorm(1 - lower_prob, mean = mu, sd = sigma)
        upper_prob <- 1 - pnorm(upper_bound, mean = mu, sd = sigma)
      } else {
        upper_bound <- obs_T
        upper_prob <- 1 - pnorm(obs_T, mean = mu, sd = sigma)
        
        # opposite-side cutoff chosen so left tail has same probability
        lower_bound <- qnorm(upper_prob, mean = mu, sd = sigma)
        lower_prob <- pnorm(lower_bound, mean = mu, sd = sigma)
      }
      
      total_tail <- lower_prob + upper_prob
      shade_left <- TRUE
      shade_right <- TRUE
      
      main_title <- bquote(
        atop(
          "Normal approximation to sampling distribution of " ~
            T["(n"[1]*","*n[2]*")"] == hat(p)[2]["(n"[2]*")"] - hat(p)[1]["(n"[1]*")"],
          "under " ~ .(hyp) ~ "," ~
            p[1] == .(p1) ~ "," ~ p[2] == .(p2) ~ "," ~
            n[1] == .(n1) ~ "," ~ n[2] == .(n2)
        )
      )
      
      sub_title <- bquote(
        "Observed " ~ T["(n"[1]*","*n[2]*")"] == .(round(obs_T, 4)) ~ "," ~
        "left tail =" ~ .(round(lower_prob, 4)) ~ "," ~
        "right tail =" ~ .(round(upper_prob, 4)) ~ "," ~
        "two-sided p-value =" ~ .(round(total_tail, 4))
      )
      
    } else {
      if (!is.null(lower_cut)) {
        lower_bound <- lower_cut
        lower_prob <- pnorm(lower_cut, mean = mu, sd = sigma)
        shade_left <- TRUE
      }
      if (!is.null(upper_cut)) {
        upper_bound <- upper_cut
        upper_prob <- 1 - pnorm(upper_cut, mean = mu, sd = sigma)
        shade_right <- TRUE
      }
      
      total_tail <- lower_prob + upper_prob
      
      main_title <- bquote(
        atop(
          "Normal approximation to sampling distribution of " ~
            T["(n"[1]*","*n[2]*")"] == hat(p)[2]["(n"[2]*")"] - hat(p)[1]["(n"[1]*")"],
          "under " ~ .(hyp) ~ "," ~
            p[1] == .(p1) ~ "," ~ p[2] == .(p2) ~ "," ~
            n[1] == .(n1) ~ "," ~ n[2] == .(n2)
        )
      )
      
      sub_title <- NULL
      if (!is.null(lower_cut) || !is.null(upper_cut)) {
        sub_title <- bquote(
          "Lower tail =" ~ .(round(lower_prob, 4)) ~ "," ~
          "Upper tail =" ~ .(round(upper_prob, 4)) ~ "," ~
          "Total p-value =" ~ .(round(total_tail, 4))
        )
      }
    }
  }
  
  if (mode == "power") {
    if (is.null(lower_cut) && is.null(upper_cut)) {
      stop("In power mode, provide at least one of lower_cut or upper_cut.")
    }
    
    if (!is.null(lower_cut)) {
      lower_bound <- lower_cut
      lower_prob <- pnorm(lower_cut, mean = mu, sd = sigma)
      shade_left <- TRUE
    }
    if (!is.null(upper_cut)) {
      upper_bound <- upper_cut
      upper_prob <- 1 - pnorm(upper_cut, mean = mu, sd = sigma)
      shade_right <- TRUE
    }
    
    total_tail <- lower_prob + upper_prob
    
    main_title <- bquote(
      atop(
        "Normal approximation to sampling distribution of " ~
          T["(n"[1]*","*n[2]*")"] == hat(p)[2]["(n"[2]*")"] - hat(p)[1]["(n"[1]*")"],
        "under the following alternative distribution," ~
          p[1] == .(p1) ~ "," ~ p[2] == .(p2) ~ "," ~
          n[1] == .(n1) ~ "," ~ n[2] == .(n2)
      )
    )
    
    sub_title <- bquote(
      "Probability to reject null =" ~ .(round(total_tail, 4)) ~
      "=" ~ "power with respect to" ~ .(hyp)
    )
  }
  
  plot(x, dens,
       type = "l",
       lwd = lwd,
       col = body_col,
       xlab = expression(hat(p)[2]["(n"[2]*")"] - hat(p)[1]["(n"[1]*")"]),
       ylab = "Density",
       main = main_title,
       sub = sub_title)
  
  if (show_tails && shade_left) {
    x_left <- x[x <= lower_bound]
    y_left <- dnorm(x_left, mean = mu, sd = sigma)
    polygon(c(min(x_left), x_left, max(x_left)),
            c(0, y_left, 0),
            col = tail_col,
            border = NA)
  }
  
  if (show_tails && shade_right) {
    x_right <- x[x >= upper_bound]
    y_right <- dnorm(x_right, mean = mu, sd = sigma)
    polygon(c(min(x_right), x_right, max(x_right)),
            c(0, y_right, 0),
            col = tail_col,
            border = NA)
  }
  
  lines(x, dens, lwd = lwd, col = body_col)
  
  invisible(list(
    x = x,
    density = dens,
    mu = mu,
    sigma = sigma,
    lower_cut = lower_bound,
    upper_cut = upper_bound,
    lower_prob = lower_prob,
    upper_prob = upper_prob,
    total = total_tail
  ))
}
```

For the following use the functions provided and solve the problem graphically.

Context: remember that checking if a pin is scammy is a very expensive operation. That's why we hire human reviewer and we sample for them with replacement several pins and they tell us if they are scammy (blue side of the cube).

## Q1: Did our scammy pins go down from last month to this month?

-   Pick the appropriate null hypothesis (negate what you're trying to prove).
-   Take $n_1 = n_2 = 10$ samples from the bags and check whether it disproves the null hypothesis $p_2>p_1$. Andrea or Van will give you numbers here.
-   Go tell your PM what you found. (summarize your findings)

## Q2: Projections

-   If it doesn't, how many more $n$'s are needed to finally reject the null, if we assume that the trend from the samples continues? (I.e. same k/n ratios)
-   Now go tell your PM about it... (summarize your findings)

## Q3: We have some idea about $p_1$ and $p_2$. Your PM wants to know how many draws to budget for.

-   Assume we suspect the truth is that $p_1 = 0.25$, and $p_2 = 0.15$.
-   Start from a certain value of n (e.g. 20)
-   Find the rejection area of null Hypothesis $p_2 > p_1$
    -   find worst case scenario $p_1$ and $p_2$ (hardest $H_0$ to disprove)
    -   find the smallest $k_1$ and/or largest $k_2$ that will still result in a rejection (smallest delta) at alpha level. That's the rejection area.
-   Calculate the probability of rejection under our suspected true scenario $p_1 = 0.3$, and $p_2 = 0.2$ (the power)
-   Go back to the beginning, do this again and adjust n until you reach power = 80%.
-   Now go tell your PM (summarize your findings)

## Q4: Do the same with the normal approximation.

## Q5: Optimal allocation of draws

-   Fix a total $n = n_1 + n_2 = 100$
-   Try different combinations of $n_1, n_2$. For each of them find the rejection area and then calculate the power. Mark down all these values that you find.
-   Which combination gave you the highest power?
-   It might be easier to do this with the normal approximation but the exact solution has more insights in it I believe.

## \[bonus\] Q6: Saving a little

-   We are sampling with replacement, therefore we risk drawing the same pin twice.
-   Can you think of a way to save a little on the human reviews?
-   On average, how much do you expect to save with that trick?

```{r}
multiplicative_factor = 10
n_1 = 50#10*multiplicative_factor
n_2 = 50#10*multiplicative_factor
k_1 = 13#0.5*multiplicative_factor
k_2 = 5#0.*multiplicative_factor

p_1_worst = (k_1+k_2)/(n_1+n_2)
p_2_worst = p_1_worst

To = k_2/n_2 - k_1/n_1


h0 <- plot_diff_prop_dist(
  n1 = n_1, n2 = n_2,
  p1 = p_1_worst, p2 = p_2_worst,
  hyp = "the null hypothesis",
  obs_T = To,
  two_tailed = TRUE,
  mode = "pvalue"
)

```

```{r}
p1_true_h1 = 0.25
p2_true_h1 = 0.15

k_1_rej_alpha = k_1
k_2_rej_alpha = k_2

T_rej_alpha = k_2_rej_alpha/n_2 - k_1_rej_alpha/n_1



h1 <- plot_diff_prop_dist(
  n1 = n_1, n2 = n_2,
  p1 = p1_true_h1, p2 = p2_true_h1,
  hyp = "the alternative hypothesis",
  lower_cut = T_rej_alpha,
  upper_cut = 2,
  mode = "power"
)

```

```{r}
plot_diff_prop_normal(
  n1 = n_1, n2 = n_2,
  p1 = p_1_worst, p2 = p_2_worst,
  hyp = "the null hypothesis",
  obs_T = To,
  two_tailed = TRUE,
  mode = "pvalue"
)
```

```{r}
plot_diff_prop_normal(
  n1 = n_1, n2 = n_2,
  p1 = p1_true_h1, p2 = p2_true_h1,
  hyp = "the alternative hypothesis",
  lower_cut = T_rej_alpha,
  upper_cut = 2,
  mode = "power"
)
```

```{r}

#| label: helper-functions
#| code-fold: true
#| code-summary: "Show helper functions"
#| 
# functions for proportions
plot_sample_prop_dist <- function(n, p,
                                  hyp = "the null hypothesis",
                                  lower_cut = NULL,
                                  upper_cut = NULL,
                                  obs_T = NULL,
                                  two_tailed = FALSE,
                                  show_tails = TRUE,
                                  mode = c("pvalue", "power"),
                                  lwd = 2,
                                  tail_col = NULL,
                                  body_col = "black",
                                  pch = 19) {
  
  mode <- match.arg(mode)
  
  if (is.null(tail_col)) {
    if (mode == "pvalue") tail_col <- "red"
    if (mode == "power")  tail_col <- "blue"
  }
  
  k <- 0:n
  phat <- k / n
  probs <- dbinom(k, size = n, prob = p)
  
  lower_tail <- rep(FALSE, length(phat))
  upper_tail <- rep(FALSE, length(phat))
  
  if (mode == "pvalue") {
    
    # ----- automatic two-tailed mode -----
    if (two_tailed) {
      if (is.null(obs_T)) {
        stop("If two_tailed = TRUE, you must provide obs_T.")
      }
      if (!(obs_T %in% phat)) {
        stop("obs_T must be one of the possible values 0, 1/n, 2/n, ..., 1.")
      }
      
      # null center
      center <- p
      
      # observed value index
      obs_idx <- which(phat == obs_T)
      
      if (obs_T <= center) {
        # observed statistic is on the left
        lower_tail <- phat <= obs_T
        lower_prob <- sum(probs[lower_tail])
        
        # on the opposite side, include the most extreme right values
        # whose total probability is <= lower_prob
        cum_prob <- 0
        upper_tail <- rep(FALSE, length(phat))
        for (i in seq_along(phat)[length(phat):1]) {
          if (phat[i] > center && cum_prob + probs[i] <= lower_prob + 1e-15) {
            upper_tail[i] <- TRUE
            cum_prob <- cum_prob + probs[i]
          }
        }
        upper_prob <- sum(probs[upper_tail])
      } else {
        # observed statistic is on the right
        upper_tail <- phat >= obs_T
        upper_prob <- sum(probs[upper_tail])
        
        # on the opposite side, include the most extreme left values
        # whose total probability is <= upper_prob
        cum_prob <- 0
        lower_tail <- rep(FALSE, length(phat))
        for (i in seq_along(phat)) {
          if (phat[i] < center && cum_prob + probs[i] <= upper_prob + 1e-15) {
            lower_tail[i] <- TRUE
            cum_prob <- cum_prob + probs[i]
          }
        }
        lower_prob <- sum(probs[lower_tail])
      }
      
      tail_region <- lower_tail | upper_tail
      total_tail <- lower_prob + upper_prob
      
      main_title <- bquote(
        atop(
          "Exact sampling distribution of " ~ T["(n)"] == hat(p)["(n)"],
          "under " ~ .(hyp) ~ "," ~ p == .(p) ~ "," ~ n == .(n)
        )
      )
      
      sub_title <- bquote(
        "Observed " ~ T["(n)"] == .(obs_T) ~ "," ~
        "left tail =" ~ .(round(lower_prob, 4)) ~ "," ~
        "right tail =" ~ .(round(upper_prob, 4)) ~ "," ~
        "two-sided p-value =" ~ .(round(total_tail, 4))
      )
      
    } else {
      # ----- ordinary manual-tail mode -----
      if (!is.null(lower_cut)) {
        lower_tail <- phat <= lower_cut
      }
      if (!is.null(upper_cut)) {
        upper_tail <- phat >= upper_cut
      }
      
      tail_region <- lower_tail | upper_tail
      
      lower_prob <- sum(probs[lower_tail])
      upper_prob <- sum(probs[upper_tail])
      total_tail <- lower_prob + upper_prob
      
      main_title <- bquote(
        atop(
          "Exact sampling distribution of " ~ T["(n)"] == hat(p)["(n)"],
          "under " ~ .(hyp) ~ "," ~ p == .(p) ~ "," ~ n == .(n)
        )
      )
      
      sub_title <- NULL
      if (!is.null(lower_cut) || !is.null(upper_cut)) {
        sub_title <- bquote(
          "Lower tail =" ~ .(round(lower_prob, 4)) ~
          "," ~ "Upper tail =" ~ .(round(upper_prob, 4)) ~
          "," ~ "Total p-value =" ~ .(round(total_tail, 4))
        )
      }
    }
  }
  
  if (mode == "power") {
    if (is.null(lower_cut)) {
      stop("In power mode, please provide lower_cut.")
    }
    
    lower_tail <- phat <= lower_cut
    tail_region <- lower_tail
    
    power_val <- sum(probs[lower_tail])
    
    main_title <- bquote(
      atop(
        "Exact sampling distribution of " ~ T["(n)"] == hat(p)["(n)"],
        "under the following alternative distribution," ~ p == .(p) ~ "," ~ n == .(n)
      )
    )
    
    sub_title <- bquote(
      "Probability to reject null =" ~ .(round(power_val, 4)) ~
      "=" ~ "power with respect to" ~ .(hyp)
    )
    
    lower_prob <- power_val
    upper_prob <- 0
    total_tail <- power_val
  }
  
  cols <- if (show_tails) {
    ifelse(tail_region, tail_col, body_col)
  } else {
    rep(body_col, length(phat))
  }
  
  plot(phat, probs,
       type = "h",
       lwd = lwd,
       col = cols,
       xlab = expression(hat(p)["(n)"]),
       ylab = "Probability",
       main = main_title,
       sub = sub_title,
       ylim = c(0, max(probs) * 1.1))
  
  points(phat, probs,
         col = cols,
         pch = pch)
  
  invisible(list(
    phat = phat,
    probs = probs,
    lower_tail = lower_tail,
    upper_tail = upper_tail,
    rejection_region = tail_region,
    lower_prob = lower_prob,
    upper_prob = upper_prob,
    total = total_tail
  ))
}
```

```{r}
plot_h0 <- plot_sample_prop_dist(
  n = 50,
  p = 0.2,
  hyp = "the null hypothesis",
  obs_T = 4/50,
  two_tailed = TRUE,
  mode = "pvalue"
)

plot_h1 <- plot_sample_prop_dist(
  n = 50,
  p = 0.15,
  hyp = "this alternative hypothesis",
  lower_cut = 4/50,
  mode = "power"
)
```